---
title: "Detecting local decoupling with lomad"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting local decoupling with lomad}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5, dpi = 96,
  message = FALSE, warning = FALSE
)
library(lomad)
```

## Overview

`lomad` (**lo**cal **m**oving **a**verage **d**ecoupling) tests whether the
underlying trends of two noisy time series are *locally similar* — moving
together up to a location shift — or whether they *decouple* over some time
windows. Given paired series following a trend-plus-noise model, the method:

1. smooths each series with a moving average of width $h$ to expose the trends;
2. computes the rolling correlation $R_t$ of the smoothed series over windows
   of length $s$;
3. compares $R_t$ pointwise against the local population correlation
   $\rho_t$ implied by a shared-trend null model, using a CLT-based $Z$
   statistic with Benjamini–Yekutieli false discovery rate control.

Intuitively, a shared trend induces the highest local correlation attainable
at the prevailing signal-to-noise ratio; genuine trend separation attenuates
$R_t$ below that benchmark. The null quantities $\rho_t$ and $V_t$ are
estimated from the data using a variogram-based noise fit that is robust to
trend contamination.

This vignette walks through the workflow twice: on simulated data where the
truth is known, and on real estuarine monitoring data.

## Simulated example

We build two scenarios from the same base trends and noise process, differing
only in the $L^2$ trend separation $d$. In the coupled scenario the two trends
are identical ($d = 0$), so the null hypothesis of local similarity holds
everywhere. In the decoupled scenario the trends separate in localized episodes
($d = 1.5$). Generating them here rather than shipping them keeps the data
transparent: every parameter of the simulation is visible.

```{r data}
make_scenario <- function(d, seed = 101) {
  tr  <- sim_trends(n = 500, d = d, method = "smooth", bw = 50, seed = seed)
  sim <- sim_noise_pair(tr, h = 5, lambda_target = 1.5, ar.coefs = 0.5,
                        seed = seed + 1)
  data.frame(t = seq_len(500), y1 = sim$y1, y2 = sim$y2,
             nu1 = tr$x1, nu2 = tr$x2)
}

cpl <- make_scenario(d = 0)      # coupled: null holds everywhere
dcp <- make_scenario(d = 1.5)    # decoupled: localized separation
str(dcp)
```

The observed series are dominated by noise; the trends (which we know here,
but would not in practice) are shown in black:

```{r raw-series, fig.height = 5.5}
# Shared y limits: the panels are only comparable on a common scale.
ylim <- range(c(dcp$y1, dcp$y2))

op <- par(mfrow = c(2, 1), mar = c(3.5, 4, 2, 1))
plot(dcp$t, dcp$y1, type = "l", col = "grey70", ylim = ylim,
     main = "Decoupled scenario: series 1", xlab = "t", ylab = "y1")
lines(dcp$t, dcp$nu1, lwd = 2)
plot(dcp$t, dcp$y2, type = "l", col = "grey70", ylim = ylim,
     main = "Decoupled scenario: series 2", xlab = "t", ylab = "y2")
lines(dcp$t, dcp$nu2, lwd = 2)
par(op)
```

The trends are close enough that the difference is hard to see at this scale —
which is the point: separation that matters for the test need not be visible by
eye. Overlaying just the two trends makes it legible:

```{r trends-overlay, fig.height = 3}
plot(dcp$t, dcp$nu1, type = "l", lwd = 2, col = "#0072B2", ylim = range(c(dcp$nu1, dcp$nu2)),
     main = "True trends", xlab = "t", ylab = expression(nu))
lines(dcp$t, dcp$nu2, lwd = 2, col = "#D55E00")
legend("topright", c(expression(nu[1]), expression(nu[2])),
       col = c("#0072B2", "#D55E00"), lwd = 2, bty = "n", horiz = TRUE)
```

### Fitting and testing

`lomad_fit()` runs the estimation pipeline and `lomad_test()` applies the
FDR-corrected pointwise test; `lomad()` wraps both. We use the smoothing
bandwidth the data were calibrated at ($h = 5$) and a rolling window of
$s = 125$ points:

```{r fit}
fit_cpl <- lomad_fit(cpl$y1, cpl$y2, h = 5, s = 125)
tst_cpl <- lomad_test(fit_cpl, alpha = 0.05)

fit_dcp <- lomad_fit(dcp$y1, dcp$y2, h = 5, s = 125)
tst_dcp <- lomad_test(fit_dcp, alpha = 0.05)
```

In the coupled scenario nothing is flagged, as it should be:

```{r coupled-result}
sum(tst_cpl$rejected, na.rm = TRUE)
```

In the decoupled scenario the test flags contiguous runs of time points:

```{r decoupled-result}
sum(tst_dcp$rejected, na.rm = TRUE)
```

`lomad_plot()` displays the smoothed series and the rolling correlation
against its null benchmark, shading the flagged windows. Two features are
worth reading carefully. The upper band is *wider* than the lower one and sits
to its left: a rejection at time $t$ concerns the whole window
$W_t = \{t-s+1, \ldots, t\}$, so the upper panel spans the smoothed values
that entered the correlation, while the lower panel marks the $t$ where $R_t$
is plotted. The grey ribbon in the lower panel is the null tolerance band,
running from $\rho_t$ down to the critical value below which $R_t$ is
flagged; the test fires exactly where $R_t$ leaves it.

```{r plot-decoupled, fig.height = 6}
lomad_plot(fit_dcp, tst_dcp)
```

Neither panel shows the test statistic itself. `lomad_test()` returns it, and
plotting it makes the decision rule concrete: $Z_t$ is standard normal under
the null, and a window is flagged when it falls below the BY-adjusted
threshold.

```{r z-stat, fig.height = 3}
plot(dcp$t, tst_dcp$Z, type = "l", col = "grey40",
     xlab = "t", ylab = expression(Z[t]), main = "Test statistic")
abline(h = 0, col = "grey70", lty = 2)
abline(h = qnorm(tst_dcp$alpha_eff), col = "steelblue", lty = 2)
rej <- which(tst_dcp$rejected)
points(rej, tst_dcp$Z[rej], col = "#D55E00", pch = 16, cex = 0.6)
```

The dashed blue line is `qnorm(tst$alpha_eff)`; orange points are the
rejections. Note it sits well below the -1.64 of an uncorrected one-sided
test at 5% — that gap is the Benjamini–Yekutieli correction paying for
`r sum(!is.na(tst_dcp$rejected))` simultaneous tests.

Because this is simulated data, we can verify the flags line up with the true
separation episodes:

```{r truth-check}
sep <- abs(dcp$nu1 - dcp$nu2)
plot(dcp$t, sep, type = "l", lwd = 2, xlab = "t",
     ylab = "|nu1 - nu2|", main = "True trend separation vs. flagged points")
flagged <- which(tst_dcp$rejected)
rug(flagged, col = "red", lwd = 2)
```

The flagged points (red rug) concentrate where the true trends actually
separate. Note the test flags windows where the *rolling correlation over the
trailing $s$ points* is depressed, so flags trail the onset of separation —
localization is at window, not point, resolution.

## Real-data example: Morro Bay

The packaged dataset `morro_bay` contains one contiguous block of
quality-controlled dissolved oxygen and pH monitoring data from the Bay Mouth
station of the Morro Bay estuary (CeNCOOS sensors), covering August to
October 2022, standardized, tidally presmoothed, and downsampled to 6-hourly
resolution (see `?morro_bay`).
Dissolved oxygen and pH are biogeochemically linked in an estuary, so
persistent local decoupling of their trends is a potential indicator of a
disturbance to the system.

```{r mb-data}
range(morro_bay$datetime)
```

With 6-hourly data we smooth over one day ($h = 4$) and use 15-day rolling
windows ($s = 60$), matching the accompanying paper's application:

```{r mb-fit}
fit_mb <- lomad_fit(morro_bay$o2, morro_bay$ph, h = 4, s = 60)
tst_mb <- lomad_test(fit_mb, alpha = 0.05)
sum(tst_mb$rejected, na.rm = TRUE)
```

```{r mb-plot, fig.height = 6}
lomad_plot(fit_mb, tst_mb, dates = morro_bay$datetime)
```

The test flags one sustained episode, in mid-September, in which the local
correlation between oxygen and pH trends falls well below its shared-trend
benchmark: $R_t$ reaches $0.03$ inside the flagged run while $\rho_t$ there
sits between 0.73 and 0.85. What the test responds to is that gap, scaled by
$\sqrt{V_t/s}$ — not the level of $R_t$ on its own. Elsewhere in the block
$R_t$ dips without being flagged, because $\rho_t$ dips with it.

## Checking the noise estimate

`lomad_fit()` estimates the noise autocovariance from residuals taken against
the *common* trend, so those residuals carry not only noise but half the gap
between the two trends, $g_t = (\nu_{1t} - \nu_{2t})/2$. The estimator is
difference-based, which suppresses that leakage but never removes it: the
lag-$l$ variogram is additive,
$V_r(l) = V_g(l) + V_Z(l)$, and $V_g(l) \geq 0$ always.

Two quick checks are worth running on real data.

**Is the noise model admissible?** The AR(1) fit uses two variogram lags, and
the identity $V(2)/V(1) = 1 + \phi$ caps that ratio at 2 for any stationary
AR(1). Equivalently $D_2 = 2V(1) - V(2)$ must be positive. When it is not, the
fitted $\hat\phi$ exceeds 1 and is clamped — a sign the residuals are outside
the model, usually because the series has been oversmoothed upstream.

**Is trend leakage distorting the estimate?** $g$ is estimable without the
common trend: smooth each series separately and halve the difference. Its
lag-1 variogram overshoots the leak by the variogram of the smoothed noise,
which the fitted parameters give you, so the difference estimates the leak.
Scaling it by $D_2$ gives a dimensionless statistic that is small when
difference-based estimation is doing its job.

```{r diagnostic}
vgram <- function(x, l) mean((x[(l + 1):length(x)] - x[1:(length(x) - l)])^2) / 2
ma    <- function(x, h) as.numeric(stats::filter(x, rep(1 / h, h), sides = 1))

# autocovariance of an h-step moving average of a process with acvf `g`
acov_smoothed <- function(g, h, kmax) {
  gl <- function(k) { k <- abs(k); if (k + 1 <= length(g)) g[k + 1] else 0 }
  vapply(0:kmax, function(k)
    sum(outer(0:(h - 1), 0:(h - 1), Vectorize(function(i, j) gl(k + i - j)))) / h^2,
    numeric(1))
}

lomad_diagnose <- function(y1, y2, fit, h) {
  keep <- which(!is.na(fit$trend))
  r1   <- y1[keep] - fit$trend[keep]
  ghat <- ((ma(y1, h) - ma(y2, h)) / 2)[keep]

  D2 <- 2 * vgram(r1, 1) - vgram(r1, 2)          # must be > 0

  # variogram of (Z1 - Z2)^(h) / 2 implied by the fitted noise parameters
  a1 <- arma_acov(fit$noise$series1$ar, fit$noise$series1$ma,
                  fit$noise$series1$sigma2, lag_max = 200)
  a2 <- arma_acov(fit$noise$series2$ar, fit$noise$series2$ma,
                  fit$noise$series2$sigma2, lag_max = 200)
  s1 <- acov_smoothed(a1, h, 1); s2 <- acov_smoothed(a2, h, 1)
  V_sn <- (s1[1] + s2[1]) / 4 - (s1[2] + s2[2]) / 4

  c(D2 = D2, S = 2 * (vgram(ghat, 1) - V_sn) / D2)
}

lomad_diagnose(morro_bay$o2, morro_bay$ph, fit_mb, h = 4)
```

`D2` is positive, so the AR(1) fit is admissible. `S` is near zero — a genuine leak can only push it positive, since $V_g$ is a mean of
squares — so trend leakage is not distorting the noise estimate here. As a
rough guide, `S` below about 0.05 indicates negligible leakage; approaching 1
means $1 - \hat\phi$ is being driven toward zero by the trend rather than by
the data, and the noise variance is badly inflated.

Both checks assume the noise model is *correctly specified*. Neither detects a
wrong model: the AR(1) fit reproduces $V(1)$ and $V(2)$ exactly by
construction, so misspecification that shows up only in the tail of the
autocovariance is invisible to them.

## Supplying your own noise model

`lomad_fit()` provides one noise estimator, a variogram-based AR(1) fit. The
inferential results themselves hold for any stationary noise process, so if
AR(1) does not describe your data you can estimate the autocovariance however
you prefer and pass it in through `noise_override`, which accepts a raw
sequence `gamma(0), gamma(1), ...` per series.

To show the mechanics, here is the *same* AR(1) estimate the package computes
internally, reproduced step by step and supplied through the override. Working
this out explicitly makes it clear where to substitute a different estimator.

```{r override}
# 1. the common trend, exactly as lomad_fit() forms it
h <- 4; s <- 60
tr <- estimate_trends(morro_bay$o2, morro_bay$ph, h = h)
keep <- which(!is.na(tr$trend))
r1 <- morro_bay$o2[keep] - tr$trend[keep]
r2 <- morro_bay$ph[keep] - tr$trend[keep]

# 2. difference-based AR(1) from two variogram lags:
#    phi = V(2)/V(1) - 1 and sigma^2 = V(2) are exact identities for AR(1)
vgram <- function(x, l) mean((x[(l + 1):length(x)] - x[1:(length(x) - l)])^2) / 2
ar1_from_variogram <- function(r) {
  V1 <- vgram(r, 1); V2 <- vgram(r, 2)
  phi <- min(max(V2 / V1 - 1, 0.01), 0.99)
  list(ar = phi, ma = numeric(0), sigma2 = V2)
}
n1 <- ar1_from_variogram(r1)
n2 <- ar1_from_variogram(r2)
c(phi1 = n1$ar, phi2 = n2$ar)

# 3. expand to autocovariance sequences and pass them in
ov <- list(series1 = list(acov = arma_acov(n1$ar, n1$ma, n1$sigma2, lag_max = 100)),
           series2 = list(acov = arma_acov(n2$ar, n2$ma, n2$sigma2, lag_max = 100)))
fit_ov <- lomad_fit(morro_bay$o2, morro_bay$ph, h = h, s = s, noise_override = ov)

# matches the built-in estimate
max(abs(fit_ov$rho - fit_mb$rho), na.rm = TRUE)
```

Replace step 2 with any estimator of your choosing — a higher-order AR fit, a
tapered sample autocovariance, or a model fitted elsewhere — and the rest of
the pipeline is unchanged. Note that estimating noise autocovariance in the
presence of an unknown trend is a genuinely hard problem: differencing removes
smooth trends but not entirely, and residual trend structure biases the
autocovariance upward. See `vignette` references and the accompanying paper's
discussion.

## Choosing h and s

- **`h` (smoothing bandwidth)** should span the dominant noise scale while
  remaining short relative to the scale on which the trends vary: larger `h`
  suppresses more noise but biases the trend estimate where it bends.
- **`s` (rolling window)** trades temporal resolution against stability of
  the local correlation; the underlying asymptotics treat `s` as large. As a
  rule of thumb, `s` should cover at least several `h`-widths.
- Passing `h = NULL, s = NULL` invokes simple length-based defaults, but for
  real applications both should be chosen from subject-matter time scales, as
  done above (1-day smoothing, 15-day windows).

## Simulating your own scenarios

The packaged data covers one coupled and one decoupled case. To ask whether the
method would see *your* problem — a separation of a given size, at your noise
level and series length — build the scenario yourself. This is also how the
accompanying paper's simulation studies are constructed.

`sim_trends()` generates a pair of trends at a controlled $L^2$ separation
$d$. How that separation is distributed over time is set by `method`:

| `method` | Separation pattern |
|---|---|
| `"dist"` | unstructured — the trends differ everywhere, with no episode structure |
| `"rate"` | separation concentrated at evenly spaced events, at rate `rate` |
| `"smooth"` | episodes at irregular times, trends pushed apart symmetrically |
| `"cross"` | as `"smooth"`, but the trends may cross rather than only diverge |

The structured methods work through a *coupling weight* $w_t$: the trends sit
on top of each other where $w_t \approx 1$ and separate as it falls. Plotting
$w_t$ alongside the trends shows exactly where the method should have
something to find.

```{r sim-structures, fig.height = 6}
op <- par(mfrow = c(3, 2), mar = c(3, 4, 2, 1))
for (m in c("rate", "smooth", "cross")) {
  args <- list(n = 500, d = 1.5, method = m, seed = 6183)
  if (m == "rate") args$rate <- 0.01 else args$bw <- 50
  tr <- do.call(sim_trends, args)

  plot(tr$x1, type = "l", col = "steelblue", lwd = 2, ylab = expression(nu),
       xlab = "", main = paste0('method = "', m, '"'),
       ylim = range(c(tr$x1, tr$x2)))
  lines(tr$x2, col = "tomato", lwd = 2)
  plot(tr$w, type = "l", col = "grey30", ylab = expression(w[t]), xlab = "t",
       main = "coupling weight")
  abline(h = 1, col = "grey70", lty = 2)
}
par(op)
```

`sim_noise_pair()` then adds ARMA noise, calibrated so the signal-to-noise
ratio hits a target at a given smoothing bandwidth — which is what makes runs
comparable across scenarios rather than depending on the trend's amplitude:

```{r sim-noise}
tr  <- sim_trends(n = 500, d = 1.5, method = "smooth", bw = 50, seed = 6183)
sim <- sim_noise_pair(tr, h = 5, lambda_target = 1.5, ar.coefs = 0.5, seed = 6184)
```

Putting it together — simulate, fit, test, and see where the method fires
relative to where the trends actually separate:

```{r sim-worked, fig.height = 4.5}
out <- lomad(sim$y1, sim$y2, h = 5, s = 125)
sum(out$test$rejected, na.rm = TRUE)

op <- par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
plot(abs(tr$x1 - tr$x2), type = "l", lwd = 2, xlab = "",
     ylab = expression(group("|", nu[1] - nu[2], "|")),
     main = "True separation, with flagged windows")
rug(which(out$test$rejected), col = "#D55E00", lwd = 2)
plot(out$fit$R, type = "l", col = "grey40", xlab = "t", ylab = "correlation")
lines(out$fit$rho, col = "steelblue", lwd = 2)
par(op)
```

Two things to expect. Flags *trail* the onset of separation, because $R_t$ is
computed over the trailing window $W_t$ — localization is at window
resolution, not point resolution. And detection depends on separation relative
to noise, not on $d$ alone: the same $d$ spread thinly over a long series is
much harder to see than the same $d$ concentrated in one episode.

## Notes

- The pointwise test is one-sided: only correlation *deficits* relative to
  the null benchmark count as evidence of decoupling.
- `lomad_test()` reports the effective significance threshold
  (`alpha_eff`) implied by the Benjamini–Yekutieli correction.
- Heed warnings from the AR(1) variogram estimator about boundary fits: a
  clamped estimate means the AR(1) model is straining, and under strong
  autocorrelation the noise variance can be overstated, which costs power.
- Oracle-style analyses (known noise parameters) can bypass estimation via
  the `noise_override` argument of `lomad_fit()`, which accepts either
  parametric coefficients or a raw autocovariance sequence (`acov`). The
  latter is the route for supplying a noise model the package does not
  itself fit.
