---
title: "Detecting local decoupling with lomad"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting local decoupling with lomad}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5, dpi = 96,
  message = FALSE, warning = FALSE
)
library(lomad)
```

## Overview

`lomad` (**lo**cal **m**oving **a**verage **d**ecoupling) tests whether the
underlying trends of two noisy time series are *locally similar* — moving
together up to a location shift — or whether they *decouple* over some time
windows. Given paired series following a trend-plus-noise model, the method:

1. smooths each series with a moving average of width $h$ to expose the trends;
2. computes the rolling correlation $R_t$ of the smoothed series over windows
   of length $s$;
3. compares $R_t$ pointwise against the local population correlation
   $\rho_t$ implied by a shared-trend null model, using a CLT-based $Z$
   statistic with Benjamini–Yekutieli false discovery rate control.

Intuitively, a shared trend induces the highest local correlation attainable
at the prevailing signal-to-noise ratio; genuine trend separation attenuates
$R_t$ below that benchmark. The null quantities $\rho_t$ and $V_t$ are
estimated from the data using a variogram-based noise fit that is robust to
trend contamination.

This vignette walks through the workflow twice: on simulated data where the
truth is known, and on real estuarine monitoring data.

## Simulated example

The packaged dataset `sim_decoupling` contains two scenarios built from the
same base trends and AR(1) noise process. In the `"coupled"` scenario the two
trends are identical ($d = 0$), so the null hypothesis of local similarity
holds everywhere. In the `"decoupled"` scenario the trends separate in
localized episodes ($d = 1.5$).

```{r data}
str(sim_decoupling)
cpl <- subset(sim_decoupling, scenario == "coupled")
dcp <- subset(sim_decoupling, scenario == "decoupled")
```

The observed series are dominated by noise; the trends (which we know here,
but would not in practice) are shown in black:

```{r raw-series, fig.height = 5.5}
op <- par(mfrow = c(2, 1), mar = c(3.5, 4, 2, 1))
plot(dcp$t, dcp$y1, type = "l", col = "grey70",
     main = "Decoupled scenario: series 1", xlab = "t", ylab = "y1")
lines(dcp$t, dcp$nu1, lwd = 2)
plot(dcp$t, dcp$y2, type = "l", col = "grey70",
     main = "Decoupled scenario: series 2", xlab = "t", ylab = "y2")
lines(dcp$t, dcp$nu2, lwd = 2)
par(op)
```

### Fitting and testing

`lomad_fit()` runs the estimation pipeline and `lomad_test()` applies the
FDR-corrected pointwise test; `lomad()` wraps both. We use the smoothing
bandwidth the data were calibrated at ($h = 5$) and a rolling window of
$s = 125$ points:

```{r fit}
fit_cpl <- lomad_fit(cpl$y1, cpl$y2, h = 5, s = 125)
tst_cpl <- lomad_test(fit_cpl, alpha = 0.05)

fit_dcp <- lomad_fit(dcp$y1, dcp$y2, h = 5, s = 125)
tst_dcp <- lomad_test(fit_dcp, alpha = 0.05)
```

In the coupled scenario nothing is flagged, as it should be:

```{r coupled-result}
sum(tst_cpl$rejected, na.rm = TRUE)
```

In the decoupled scenario the test flags contiguous runs of time points:

```{r decoupled-result}
sum(tst_dcp$rejected, na.rm = TRUE)
```

`lomad_plot()` displays the smoothed series and the rolling correlation
against its null benchmark, shading the flagged windows:

```{r plot-decoupled, fig.height = 6}
lomad_plot(fit_dcp, tst_dcp)
```

Because this is simulated data, we can verify the flags line up with the true
separation episodes:

```{r truth-check}
sep <- abs(dcp$nu1 - dcp$nu2)
plot(dcp$t, sep, type = "l", lwd = 2, xlab = "t",
     ylab = "|nu1 - nu2|", main = "True trend separation vs. flagged points")
flagged <- which(tst_dcp$rejected)
rug(flagged, col = "red", lwd = 2)
```

The flagged points (red rug) concentrate where the true trends actually
separate. Note the test flags windows where the *rolling correlation over the
trailing $s$ points* is depressed, so flags trail the onset of separation —
localization is at window, not point, resolution.

## Real-data example: Morro Bay

The packaged dataset `morro_bay` contains two contiguous blocks of
quality-controlled dissolved oxygen and pH monitoring data from the Bay South
station of the Morro Bay estuary (CeNCOOS sensors), standardized, tidally
presmoothed, and downsampled to 6-hourly resolution (see `?morro_bay`).
Dissolved oxygen and pH are biogeochemically linked in an estuary, so
persistent local decoupling of their trends is a potential indicator of a
disturbance to the system.

```{r mb-data}
b2 <- subset(morro_bay, block == 2)     # Apr-Dec 2020
range(b2$datetime)
```

With 6-hourly data we smooth over one day ($h = 4$) and use 15-day rolling
windows ($s = 60$), matching the accompanying paper's application. The noise
in this system has non-trivial autocorrelation structure, so we use the
BIC-selected AR noise model rather than the AR(1) default:

```{r mb-fit}
fit_b2 <- lomad_fit(b2$o2, b2$ph, h = 4, s = 60, noise_method = "arma")
tst_b2 <- lomad_test(fit_b2, alpha = 0.05)
sum(tst_b2$rejected, na.rm = TRUE)
```

```{r mb-plot, fig.height = 6}
lomad_plot(fit_b2, tst_b2, dates = b2$datetime)
```

The test flags one sustained episode in which the local correlation between
oxygen and pH trends falls significantly below its shared-trend benchmark.
For comparison, the shorter autumn 2023 block shows no flagged points — the
two trends remain coupled throughout:

```{r mb-block21}
b21 <- subset(morro_bay, block == 21)
fit_b21 <- lomad_fit(b21$o2, b21$ph, h = 4, s = 60, noise_method = "arma")
tst_b21 <- lomad_test(fit_b21, alpha = 0.05)
sum(tst_b21$rejected, na.rm = TRUE)
```

## Choosing h and s

- **`h` (smoothing bandwidth)** should span the dominant noise scale while
  remaining short relative to the scale on which the trends vary: larger `h`
  suppresses more noise but biases the trend estimate where it bends.
- **`s` (rolling window)** trades temporal resolution against stability of
  the local correlation; the underlying asymptotics treat `s` as large. As a
  rule of thumb, `s` should cover at least several `h`-widths.
- Passing `h = NULL, s = NULL` invokes simple length-based defaults, but for
  real applications both should be chosen from subject-matter time scales, as
  done above (1-day smoothing, 15-day windows).

## Notes

- The pointwise test is one-sided: only correlation *deficits* relative to
  the null benchmark count as evidence of decoupling.
- `lomad_test()` reports the effective significance threshold
  (`alpha_eff`) implied by the Benjamini–Yekutieli correction.
- For noise with complex or oscillatory autocorrelation, prefer
  `noise_method = "arma"`, and heed warnings from the AR(1) variogram
  estimator about boundary fits.
- Oracle-style analyses (known noise parameters) can bypass estimation via
  the `noise_override` argument of `lomad_fit()`.
